\chapter{Introduction}

\section{Motivation}

The quadratic time and memory complexity of self-attention
\cite{vaswani2017attention} has long defined the practical ceiling
of sequence modelling. As context windows expand toward millions
of tokens, this ceiling translates into prohibitive deployment
costs in memory, latency, and energy. The field has consequently
shifted toward sub-quadratic alternatives, and linear-time
recurrent models, in particular State Space Models (SSMs) and
the RWKV lineage, have moved from academic curiosities into a
production substrate for frontier-scale systems
\cite{li2025minimax01,liu2025hunyuanturbos}.

This industrial adoption is evidenced by several recent
deployments. MiniMax-01 demonstrated that linear attention can
drive a $456$\,B-parameter mixture-of-experts model with context
windows reaching $4$\,M tokens \cite{li2025minimax01}. Tencent's
Hunyuan-TurboS \cite{liu2025hunyuanturbos} and NVIDIA's
Nemotron-H family \cite{blakeman2025nemotronh} replace the
majority of self-attention layers with Mamba blocks, reporting up
to a threefold inference speed-up at parity. Further examples
include AI21's Jamba \cite{lieber2024jamba}, the Zamba2 suite
\cite{glorioso2024zamba2}, and the Evo~2 genomic foundation model,
which uses a linear convolutional multi-hybrid to reach a $1$\,M
nucleotide context \cite{brixi2026evo2}. Now that these models
form a production substrate, their structural limitations
translate directly into measurable deployment costs.

Underneath this architectural variety, the models share a
recurrent core and linear-time token mixing. They also share
characteristic failure modes: an inability to resolve local
structure at the correct scale, unreliable associative recall,
and restricted recurrent dynamics. Numerous fixes have been
proposed, including quad-directional spatial shifts in vision,
depthwise convolutions in audio, and decorrelation steps in
language modelling, yet they remain fragmented. Most improvements
are evaluated in isolation and rarely cross-checked against
different linear backbones. This thesis addresses the missing
cross-cut by treating \textit{mechanism-prior alignment} as a
hypothesis: the gain a mechanism delivers should be predictable
from how much of the targeted deficit the underlying backbone
leaves uncovered.

\section{Contributions}

We introduce three functional mechanisms that address documented
expressivity deficits of linear-time recurrent models. The
mechanisms target short-range temporal hierarchy through learned
local context, reduce interference in associative recall via
feature decorrelation, and expand the recurrent transition class
by lifting the state update beyond real-diagonal constraints.
Together they enable a controlled investigation of the
interaction between architectural structure and computational
capacity.

To evaluate these mechanisms we employ a cross-architecture
matrix across multiple linear-time backbones, using specialised
probes that isolate specific failure modes at several parameter
scales under a unified training regime. This systematic
evaluation moves beyond aggregate benchmark numbers and observes
the transferability of gains directly across architectural
primitives.

The resulting evidence supports the central claim of the thesis:
mechanism gains are proportional to architectural deficits.
Performance improvements are not arbitrary; they depend on
whether a mechanism addresses a specific gap left uncovered by
the base model. The claim is reinforced by both productive
improvements and informative null results, in particular by
cases where a mechanism is engaged but unhelpful, or absorbed by
an existing native primitive, validating mechanism-prior
alignment as a predictive criterion.

\section{Structure Of The Thesis}
