\chapter{Experimental Setup}\label{app:experimental-setup}

This appendix records the configuration that produced every empirical
result reported in Chapter~\ref{ch:experiments}. The content
complements the chapter-level methodology with the per-run
hyperparameter schedule, the ASR baseline architecture, the
dataset-construction details, the MQAR cohort configuration, and the
hardware environment. The implementations of the proposed mechanisms
and the experimental harness are available upon request from the
author.

\section{Training configuration}\label{app:setup:training}

The training schedule is the locked 50-epoch matched budget shared by
every LibriSpeech and Common Voice cell of the matrix. The values in
Table~\ref{tab:setup:hyperparams} apply to every reported cell at
both the 7M and 14M scales unless otherwise noted.

\begin{table}[!htbp]
\centering
\small
\caption[Locked training hyperparameters]{Training hyperparameters
shared by every LibriSpeech and Common Voice cell. The schedule is
locked across architectures and scales.}
\label{tab:setup:hyperparams}
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{}p{4.5cm} X@{}}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Optimiser & AdamW \\
Peak learning rate & $3 \times 10^{-4}$ \\
Weight decay & $0.01$ \\
Learning-rate schedule & Cosine decay with $1000$-step linear warmup \\
Epochs (matched budget) & $50$ \\
Early-stopping patience & $15$ epochs on dev CER \\
Batch budget & $\le 300$~s of audio per batch \\
Gradient clipping ($\ell_{2}$ norm) & $5.0$ \\
SpecAugment policy & LD: freq mask $F = 27$, time mask $T = 100$ \\
SpecAugment masks per utterance & $2$ frequency masks, $2$ time masks \\
Dropout & $0.1$ \\
Random seed & $42$ \\
Mamba-2 SSD chunk size & $64$ \\
\bottomrule
\end{tabularx}
\end{table}

The cosine schedule is applied to the learning rate after the
$1000$-step linear warmup completes; the schedule decays to a small
positive floor over the remaining steps of the 50-epoch budget. The
patience-15 early-stopping rule is set against dev CER and was not
triggered for any reported cell, since the matched 50-epoch budget
sits within the convergence horizon of all three backbones at both
scales.

The batch budget is duration-based rather than utterance-based.
Each training mini-batch packs utterances until the cumulative
duration reaches $300$~s, which keeps GPU occupancy stable across
the short-to-medium audio length distribution of LibriSpeech and
Common Voice without requiring per-step length statistics.
Validation batches use a fixed duration limit of $300$~s as well,
with no SpecAugment.

The single-seed convention is the matrix-wide default. The
matrix-as-artefact framing of Chapter~\ref{ch:experiments} relies
on the single coherent direction across approximately seventy
statistical tests rather than on any single-cell multi-seed claim.

\section{ASR baseline architecture}\label{app:setup:arch}

The ASR model is identical across cells except for the time-mixing
backbone. Table~\ref{tab:setup:arch} summarises the shared pipeline
from audio frontend to CTC head.

\begin{table}[!htbp]
\centering
\small
\caption[ASR baseline architecture]{ASR baseline pipeline shared by
every reported cell. The backbone row is the only stage that varies
across cells.}
\label{tab:setup:arch}
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{}p{3.5cm} X@{}}
\toprule
\textbf{Stage} & \textbf{Configuration} \\
\midrule
Audio frontend & Log-mel spectrogram, $80$ mel bins, $25$~ms window,
$10$~ms hop, sample rate $16$~kHz \\
\addlinespace
Subsampling & $2\times$ acoustic subsampling before the encoder
backbone, applied via a small convolutional stack \\
\addlinespace
Encoder block (per layer) & Time-mix block (backbone-specific) plus a
channel-mix block, each with residual connection and pre-LayerNorm \\
\addlinespace
Backbone & Per-cell choice of causal RWKV-6, Mamba-2, or Linear
Attention; or the LION bidirectional adaptation of the same three \\
\addlinespace
CTC head & Linear projection to vocabulary followed by softmax and
the connectionist temporal classification
loss~\cite{graves2006ctc} \\
\addlinespace
Decoding & Greedy CTC, no language model rescoring, no beam search \\
\bottomrule
\end{tabularx}
\end{table}

The encoder configuration at the 7M scale is $d_{\text{model}} = 256$,
$L = 6$ layers, $H = 4$ heads, head dimension $64$. At the 14M
scale the layer count doubles to $L = 12$ with the other widths
unchanged. The Mamba-2 backbone uses state dimension
$d_{\text{state}} = 64$ and SSD chunk size $T_{c} = 64$ throughout.
Total parameter counts per cell range from $6.26$~M for vanilla
Linear Attention at 7M to $13.79$~M for the RWKV-6 DHO cell at
14M; per-cell counts are reported in
Table~\ref{tab:results:master-matrix-causal-7m} of
Appendix~\ref{app:results}.

The LION wrapper preserves the time-mix and channel-mix
abstraction~\cite{afzal2025linear}, replacing the causal scan kernel
with a parallel $T \times T$ attention computation under a symmetric
decay mask. The decay-class mapping per backbone is described in
Section~\ref{sec:lion} and Section~\ref{sec:proposed:lion}; the
controlled LION-S deployment on Linear Attention introduces a
per-token sigmoid decay that is not present in the underlying
causal backbone, and is used as the falsification control for the
decay-as-prerequisite finding of
Chapter~\ref{ch:experiments}.

\section{Datasets}\label{app:setup:datasets}

Three datasets carry the empirical evaluation, chosen for the
expressivity axis each one exercises rather than for aggregate-score
competitiveness.

\subsection{LibriSpeech clean-100}\label{app:setup:librispeech}

The principal probe is the train-clean-100 subset of
LibriSpeech~\cite{panayotov2015librispeech}: $100$ hours of
audiobook-source read English speech recorded at $16$~kHz, with the
official train, dev-clean, and test-clean splits used unchanged.
The character-level vocabulary contains the $26$ lowercase letters,
the apostrophe, the space character, and a dedicated CTC blank
token, for a head dimension of $29$.

Audio is preprocessed with the parameters of
Table~\ref{tab:setup:arch}: $80$-mel log-spectrograms with $25$~ms
window and $10$~ms hop, followed by the $2\times$ acoustic
subsampling described above. Utterances longer than $20$~s and
shorter than $0.5$~s are filtered from the training stream;
SpecAugment LD is applied during training as specified in
Section~\ref{app:setup:training} and disabled during evaluation.
Test evaluation uses full-utterance greedy CTC decoding with no
language model rescoring.

\subsection{Common Voice English 100h}\label{app:setup:cv}

The cross-distribution probe is a $100$-hour balanced subset of
the Common Voice English corpus~\cite{ardila2019commonvoice}. The
dataset is selected for its broader acoustic variance relative to
LibriSpeech. Common Voice is community-collected from several
thousand contributors, with explicit accent labels across
regional varieties of English and recording-condition
heterogeneity arising from non-studio crowd-sourced contribution.
The audiobook-source LibriSpeech recording protocol does not
expose comparable variability along these axes. The thesis
pre-registers this distinction as a probe property prior to any
experiment on the proposed mechanisms.

The training subset is constructed from the validated split of
the Common Voice English release. The validated split aggregates
utterances that have received sufficient up-votes from the Common
Voice review pipeline; its total duration varies by release
version and is in the multi-thousand-hour range at recent
releases. The selection protocol of this thesis proceeds as
follows. The validated split is partitioned by speaker
identifier; speakers are drawn uniformly without replacement
under the matrix-wide random seed ($42$,
Table~\ref{tab:setup:hyperparams}); utterances from each drawn
speaker are accumulated until the cumulative duration crosses the
LibriSpeech train-clean-100 budget within rounding. Inclusion
criteria match the LibriSpeech pipeline: utterances longer than
$20$~s and shorter than $0.5$~s are filtered. Validated-split
utterances without an associated speaker identifier are excluded
from the speaker-stratified sampling, since the no-overlap
guarantee at evaluation time relies on a deterministic speaker
key. The dev and test partitions are the released Common Voice
splits restricted to the sampled speaker set, which preserves the
no-speaker-overlap property at evaluation time. Audio
preprocessing matches the LibriSpeech pipeline; the only change
is the vocabulary, which retains the same character set since
both corpora are English.

The pre-experimental contrast between the two probes is recorded
in Table~\ref{tab:librispeech-vs-cv} of
Section~\ref{sec:experiments:framework}. The numerical anchors of
that table reference the source publications: LibriSpeech
train-clean-100 carries $251$ speakers across $100.6$~h of
audiobook-source read English~\cite{panayotov2015librispeech}, and
Common Voice English aggregates volunteer contributions across
several thousand speakers in its validated split with explicit
accent labels across regional English
varieties~\cite{ardila2019commonvoice}. The $100$~h Common Voice
subset of this thesis is sampled from the validated split under
the stratification protocol described above.

\subsection{Multi-Query Associative Recall}\label{app:setup:mqar}

The synthetic axis-2 probe is the multi-query associative recall
benchmark of~\citet{arora2023zoology}. Each example is a sequence
of $(k, v)$ pairs followed by query keys whose values must be
retrieved. Sequence length $T$ sweeps the canonical $\{64, 256,
1024\}$ values; the number of stored pairs $K$ scales with $T$ such
that the difficulty parameter $K / d$ stays meaningful at every
length: for $T = 64$, $K = 16$; for $T = 256$, $K = 64$; for
$T = 1024$, $K = 256$. The vocabulary contains $8192$ key tokens
and $8192$ value tokens drawn from disjoint ranges. Training uses
unlimited synthetic data with cross-entropy loss on the
value-prediction positions; the success criterion is per-sequence
accuracy reaching $0.9$. Each MQAR cell is run for up to $30\,000$
steps with early stopping when per-sequence accuracy reaches
$0.999$ on a held-out validation batch.

The thirteen-backbone cohort uses a shared encoder
configuration; only the time-mixing block varies between cells.
The shared configuration is two encoder layers with
$d_{\text{model}} = 64$ and the architecture-specific time-mix
substituted in. The causal Transformer baseline is
parameter-matched to the linear-time backbones at the same
configuration, with standard multi-head attention substituted in
place of the linear-time time-mix; the head count is $H = 1$ at
head dimension $64$, the channel-mix block expands by a factor
of $4$ to a feed-forward dimension of $256$, and no positional
encoding is applied beyond the implicit sequence-position signal
of standard causal attention. The fair-comparison interpretation
of the $T = 1024$ FAIL of the causal Transformer therefore
attributes to the architecture rather than to a parameter-budget
or capacity gap. The optimisation schedule is shared across
every cohort cell: AdamW with peak learning rate
$3 \times 10^{-4}$ and a $1000$-step linear warmup followed by
cosine decay to a small positive floor, matching the LibriSpeech
schedule of Section~\ref{app:setup:training} as closely as the
step-based MQAR budget allows.

\section{Hardware environment}\label{app:setup:hw}

All cells were trained on a single workstation equipped with NVIDIA
RTX PRO 6000 Blackwell GPUs ($48$~GB of VRAM each). Single-GPU
training is the matrix-wide default. The only exception is the LA
LION-S DHO cell at the 7M scale, which used 4-GPU data-parallel
distributed training to fit the bidirectional complex-attention
scan within a practical wall-clock window; the gradient
accumulation pattern was set to keep the effective batch budget at
the matrix-wide $300$~s of audio. The software stack consists of
PyTorch with CUDA acceleration; the Mamba-2 backbone uses the
official SSD CUDA kernel from the
\texttt{state-spaces/mamba} reference
repository~\cite{gu2024mambacode}, and the RWKV-6 backbone uses
the reference time-mix and channel-mix block implementations
from the RWKV Foundation~\cite{rwkv2024block}.
