\chapter{Mechanism Implementation Details}\label{app:mechanisms}

This appendix records the canonical implementation parameters for the
three mechanisms presented in Chapter~\ref{ch:proposed}. The contents
parallel the chapter structure: a Multi-Scale Depthwise Convolution
(MSDC) section, a Damped Harmonic Oscillator (DHO) section, a Chunked
Value Decorrelation (CVD) section, and a unified LION wrapper section.
Each section identifies the codebase modules where the mechanism is
implemented, lists the constructor parameters and initialisation
schedule, and documents any architecture-specific deployment detail
that the chapter abstracted from. Equations referenced as
``Chapter~\ref{ch:proposed}'' carry the same numbering as in the
proposed-solution chapter.

\section{Multi-Scale Depthwise Convolution}\label{app:mechanisms:msdc}

\subsection{Module location and constructor}

MSDC is implemented as a single class
\texttt{MultiDilationDWConvShift} in
\texttt{experiments/formal\_v1/src/models/conv\_shift.py}. The
class is constructed at backbone-build time and inserted at the
input side of every encoder layer. The constructor parameters are
listed in Table~\ref{tab:msdc-params}.

\begin{table}[h]
\centering
\small
\caption[MSDC constructor parameters]{Constructor parameters of the
Multi-Scale Depthwise Convolution mechanism. The values are shared
across the three architectures and across the 7M and 14M scales.}
\label{tab:msdc-params}
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.10}
\begin{tabularx}{\linewidth}{@{}p{4.5cm} X@{}}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Kernel size $k$ & $3$ \\
Dilation set $\mathcal{D}$ & $\{1, 2, 4, 8\}$ \\
Padding mode & Symmetric (zero-padded on both sides) \\
Convolution structure & Depthwise (channel-separable) \\
Mixing weights $\alpha_{d}$ & Per-layer, per-channel,
$\alpha_{d} \in \mathbb{R}^{D}$ \\
Insertion point & Input side, before the layer's time-mix block \\
\bottomrule
\end{tabularx}
\end{table}

For each layer $\ell$, the mechanism implements
$\mathrm{MSDC}(x)[t] = \sum_{d \in \mathcal{D}} \alpha_{d}^{(\ell)}
\odot \mathrm{DWConv}_{k, d}(x)[t]$ as in
equation~(\ref{eq:msdc-generic}) of Chapter~\ref{ch:proposed}. The
four dilated branches share the input tensor and run as parallel
depthwise convolutions; their outputs are mixed by the per-channel
$\alpha_{d}^{(\ell)}$ weights and the result becomes the input to the
backbone time-mix block via a pre-block residual.

\subsection{Initialisation}\label{app:mechanisms:msdc:init}

The mechanism initialisation is the load-bearing piece. The
constructor sets $\alpha_{d=1}^{(\ell)} = 1.0$ in every channel of
every layer. For dilations $d > 1$, the initialisation is
$\alpha_{d > 1}^{(\ell)} = 0.01$ in every channel. The branch-1
convolution kernel is initialised to the centre tap, weights
$[0.5, 0, 0.5]$ and zero bias, so that
$\mathrm{DWConv}_{k=3, d=1}(x)[t]$ at initialisation reduces to a
fixed two-tap symmetric average; for $d > 1$ the branch convolution
weights are drawn from $\mathcal{N}(0, (0.01)^{2})$.

The non-zero initialisation of $\alpha_{d > 1}$ and of the branch
weights for $d > 1$ is required for the dilated branches to receive
gradient. With both legs at zero, the gradient of $\alpha_{d}$ is
proportional to the branch output (which is zero) and the gradient of
the branch weights is proportional to $\alpha_{d}$ (which is zero), so
the dilated branches remain frozen for the entirety of training and
the mechanism collapses to a single-dilation depthwise convolution.
The $10^{-2}$ value perturbs one leg of the product enough for SGD to
escape the flat region while leaving the initial output close enough
to the unperturbed centre tap that the zero-regression contract at
initialisation is preserved within activation noise.

\subsection{Per-architecture insertion}

The insertion point of MSDC is structurally consistent across the
three backbones, preserving the architecture-agnostic generic form of
Chapter~\ref{ch:proposed}. On RWKV-6, MSDC is added in front of the
Time-Mix block as part of the per-layer token-shift residual. On
Mamba-2, MSDC is added on the input side of the SSD block. On
Linear Attention, MSDC is added in front of the per-layer
projection that produces the kernel feature map. The same
\texttt{MultiDilationDWConvShift} instance is used in every
backbone; the dispatch lives in the per-backbone encoder constructor
under \texttt{experiments/formal\_v1/src/models/}.

\section{Damped Harmonic Oscillator}\label{app:mechanisms:dho}

\subsection{Module location}

DHO is implemented as a per-backbone fork of the time-mix block. The
canonical entry points are
\texttt{experiments/formal\_v1/src/models/rwkv6\_time\_mix.py}
(RWKV-6),
\texttt{experiments/formal\_v1/src/models/mamba2\_rse.py} (Mamba-2),
and
\texttt{experiments/formal\_v1/src/models/linear\_attn\_rse.py}
(Linear Attention). The three implementations share the same
mathematical form, equation~(\ref{eq:dho-generic}) of
Chapter~\ref{ch:proposed}, and differ only in how the resulting
block-complex transition is composed with the backbone's native scan
kernel.

\subsection{Block-complex transition}

The recurrent state is partitioned into $K/2$ pairs of channels per
head, and the transition operator on each pair is the rotation-and-
damping operator
\[
G_{t, h, b} = e^{-\lambda_{\mathrm{eff}, t, h, b}} \cdot R(\theta_{t, h, b}),
\quad
R(\theta) = \begin{pmatrix} \cos\theta & -\sin\theta \\ \sin\theta & \phantom{-}\cos\theta \end{pmatrix},
\]
with $h$ indexing heads and $b \in \{0, \ldots, K/2 - 1\}$ indexing
the per-head block. The angle $\theta_{t, h, b}$ is data-dependent
and is produced by a low-rank LoRA projection of the per-token input.

\subsection{Depth-graded $\theta$-clip schedule}

The angle is bounded by a depth-graded clip schedule that grows from
a tight constraint in early layers to a wider budget in deep layers.
At the 7M scale ($L = 6$ layers), the schedule is
\[
\theta_{\max}(\ell) =
\bigl(\pi/8,\ \pi/8,\ \pi/4,\ \pi/4,\ \pi/2,\ \pi/2\bigr),
\quad \ell \in \{0, 1, 2, 3, 4, 5\}.
\]
At the 14M scale ($L = 12$ layers), the schedule extends in lockstep
by repeating each pair: layers $0$ to $3$ at $\pi/8$, layers $4$ to
$7$ at $\pi/4$, layers $8$ to $11$ at $\pi/2$. The schedule is
fixed; it is not learned, and there is no within-matrix variation
in its values.

\subsection{LoRA rank schedule}

The data-dependent rotation angle $\theta_{t, h, b}$ at layer $\ell$
is produced by a rank-$r(\ell)$ LoRA projection of the per-token
input, with the rank schedule rising in lockstep with the
$\theta$-clip schedule. At the 7M scale,
\[
r(\ell) = (16, 16, 32, 32, 48, 48), \quad \ell \in \{0, 1, 2, 3, 4, 5\},
\]
and at the 14M scale the schedule extends by pair repetition to
$(16, 16, 16, 16, 32, 32, 32, 32, 48, 48, 48, 48)$. The
deeper layers, which receive the wider rotation budget, also receive
the higher-capacity projection.

\subsection{Viscosity coefficient}

The Rayleigh viscosity coefficient $\eta_{h, b}$ is a learnable
scalar of shape $(\text{n\_heads}, \text{n\_blocks})$, zero-initialised.
The effective decay used by the transition is
\[
\lambda_{\mathrm{eff}, t, h, b} = \lambda_{\mathrm{raw}, h, b} +
\eta_{h, b} \cdot \theta_{t, h, b}^{2}.
\]
The zero-initialisation makes DHO reduce exactly to plain
block-complex rotation at training step zero, so the mechanism
satisfies the zero-regression-at-init contract. The viscosity term
engages only as $\eta_{h, b}$ moves under the gradient.

\subsection{Per-architecture deployment}

On RWKV-6, the block-complex transition replaces the per-channel
data-dependent diagonal decay $w_{t}$ on the time-mix path. The
implementation packs each pair of state channels into a single
complex scalar and runs a chunked complex-arithmetic scan within
chunks of $T_{c} = 64$, with inter-chunk state carried serially. On
Mamba-2, the block-complex transition is composed with the SSD
selective-decay continuous-time form: the scalar identity
$a_{t} I$ of the dual form is wrapped in a per-pair rotation-and-
damping block, yielding a block-diagonal closure of pairs of state
channels around the original $a_{t}$. On Linear Attention, where the
vanilla transition is the identity, the block-complex operator
introduces an entirely new transition class: the recurrent state
inherits a non-trivial rotation-and-damping operator that is absent
in the vanilla causal LA backbone.

\section{Chunked Value Decorrelation}\label{app:mechanisms:cvd}

\subsection{Module locations}

CVD has three architecture-specific deployments. The dispatch points
are:
\begin{itemize}
\item \texttt{rwkv6\_time\_mix.\_apply\_lucid\_recurrent} for the
chunked recurrent form on RWKV-6;
\item \texttt{mamba2\_kernels.\_apply\_lucid\_mamba2\_chunked} for
the SSD chunk-local form on Mamba-2;
\item \texttt{lion\_attention.lion\_attention\_with\_lucid} for the
parallel form used by the LION encoder, alongside the causal LA
deployment that uses the same in-block preconditioner construction.
\end{itemize}
The three deployments share the preconditioner equation of
Chapter~\ref{ch:proposed}, equation~(\ref{eq:cvd-generic}), and
differ in the substrate on which $\tilde V$ is consumed.

\subsection{Preconditioner equation}

For a chunk of keys $K_{c} \in \mathbb{R}^{T_{c} \times d}$, the
mechanism builds the within-chunk preconditioner
\[
P_{c} = \exp\!\left(
\tau_{h} \cdot \left(
\frac{K_{c}^{\mathrm{RN}} (K_{c}^{\mathrm{RN}})^{\top}}{\sqrt{d}} - \sqrt{d}
\right)
\right) + \varepsilon \, I,
\quad
K_{c}^{\mathrm{RN}} = \sqrt{d} \cdot \frac{K_{c}}{\|K_{c}\|_{2}},
\]
solves $P_{c} \tilde V_{c} = V_{c}$ once per chunk per head, and
feeds $\tilde V_{c}$ into the unchanged scan of the underlying
backbone. The $-\sqrt{d}$ shift makes $P_{c}$ unit-diagonal at
initialisation in the RMS-normalised key construction, which prevents
each token from subtracting its own value at readout.

\subsection{Per-head learnable temperature}

The temperature $\tau_{h}$ is per-head and learnable, parameterised
through softplus of a raw parameter $\tau_{\mathrm{raw}, h}$:
\[
\tau_{h} = \mathrm{softplus}(\tau_{\mathrm{raw}, h}).
\]
The initialisation of $\tau_{\mathrm{raw}, h}$ is architecture-split.
On RWKV-6, on Mamba-2, and on the LA-LION (LION-LIT and LION-S)
deployments, the initialisation is $\tau_{\mathrm{raw}, h} =
\log(e - 1)$, which gives $\tau_{h} = 1.0$ at training step zero. On
the causal LA deployment, the initialisation is
$\tau_{\mathrm{raw}, h} = 0$, which gives
$\tau_{h} = \mathrm{softplus}(0) = \ln 2 \approx 0.693$ at training
step zero. The discrepancy reflects the historical sequence in which
the deployments were wired, with the causal LA path preceding the
unification of the initialisation schedule across the other paths;
the trained values of $\tau_{h}$ overlap across the four
deployments after a small number of training epochs and the
discrepancy does not materially affect the reported numbers.

\subsection{Architecture-split regulariser}

The regulariser $\varepsilon \, I$ is added to keep the chunk-local
linear solve well-conditioned at trained temperatures. The value is
architecture-split:
\[
\varepsilon =
\begin{cases}
10^{-4} & \text{on the Mamba-2 deployment,} \\
10^{-6} & \text{on the RWKV-6 and Linear Attention deployments.}
\end{cases}
\]
The Mamba-2 value is larger because the SSD chunk-local Gram matrix,
constructed from highly correlated B-rows under trained Mamba-2
weights, becomes ill-conditioned at trained temperatures around
$\tau_{h} \approx 1.5$; the smaller RWKV-6 and LA value is paired
with an element-wise singular-fallback path through
\texttt{torch.linalg.solve\_ex} that handles the rare per-chunk
near-singular case without aborting the run.

\subsection{Chunk size}

The chunk size used by the preconditioner construction is
$T_{c} = 64$ across all three backbones. On RWKV-6, the chunk size
matches the chunked WKV scan. On Mamba-2, the chunk size matches the
SSD dual-form chunk size, which is also $64$ in the locked
configuration. On the LION parallel form, the chunk size of the
preconditioner block is also $64$, applied within the bidirectional
parallel attention rather than across full sequence length.

\subsection{Mamba-2 deployment: C-correlation analog}

The Mamba-2 deployment of CVD uses the chunk-local
$\mathbf{C}_{c}$ tensor, the query analog in the SSD scalar-identity
dual form, as the correlation source rather than the
$\mathbf{B}_{c}$ tensor. The chunk-local Gram matrix is built as
$\mathbf{C}_{c}^{\mathrm{RN}} (\mathbf{C}_{c}^{\mathrm{RN}})^{\top}$
and the preconditioner construction proceeds as in
equation~(\ref{eq:cvd-generic}). The choice of $\mathbf{C}$ rather
than $\mathbf{B}$ pairs the algebra with the prose framing of
key-similarity decorrelation in Chapter~\ref{ch:proposed} and
produces marginally cleaner numbers in the experimental matrix; both
correlation sources transfer productively, with test CER differing
by less than $10^{-3}$ on the Mamba-2 backbone.

\subsection{Relationship to LUCID}

The relationship between CVD and LUCID
Attention~\cite{duvvuri2026lucid} is shared logical motivation, not
direct adaptation. LUCID operates on the full $T \times T$
softmax-attention matrix inside a quadratic-complexity Transformer,
preserves $O(N^{2} d)$ complexity, and combines the preconditioner
with the softmax weights in front of the value path. CVD operates
inside the recurrence of a linear-time RNN, never materialises the
full $T \times T$ matrix, uses chunk-local $T_{c} \times T_{c}$
systems exclusively, introduces a per-head learnable temperature
$\tau_{h}$ in place of LUCID's fixed $1/\sqrt{d}$ scaling, and adds
the $\varepsilon I$ regulariser absent from LUCID. The shared move
is the use of a key-similarity preconditioner to decorrelate values
before they enter the memory operation, motivated by the same
observation that correlated keys cause readout interference in any
associative-memory implementation. This thesis contributes the
formulation of value decorrelation on linear-time recurrent
substrates as a class, three architecture-specific deployments of
that class (RWKV-6 chunked recurrent, Mamba-2 SSD-chunk-local,
LION parallel), and the empirical demonstration of
asymmetric transfer across the family.

\section{Unified LION wrapper}\label{app:mechanisms:lion}

\subsection{Realisation as a shared interface convention}

The LION wrapper is realised as a shared interface convention rather
than a single shared class, because each backbone's parallel form
needs its own kernel implementation. The three backbone-specific
LION dispatches are summarised in Table~\ref{tab:lion-dispatch}.

\begin{table}[h]
\centering
\small
\caption[LION wrapper dispatch per backbone]{Per-backbone realisation
of the LION bidirectional wrapper. The three deployments share the
parallel form and the symmetric-mask attention semantics of
\citet{afzal2025linear}, while running through backbone-specific
kernel implementations.}
\label{tab:lion-dispatch}
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.10}
\begin{tabularx}{\linewidth}{@{}p{3.0cm} p{3.5cm} X@{}}
\toprule
\textbf{Backbone} & \textbf{Mode parameter} & \textbf{Dispatched kernel} \\
\midrule
RWKV-6 &
\texttt{RWKV6TimeMix(mode=lion)} &
\texttt{lion\_attention.lion\_parallel\_attention} \\
\addlinespace
Mamba-2 &
\texttt{Mamba2Block(mode=lion)} &
\texttt{mamba2\_kernels.ssd\_scan\_lion} \\
\addlinespace
Linear Attention &
\texttt{LIONLinearAttentionEncoder(decay\_mode \(\in\) \{lit, s\})} &
\texttt{LIONLinearAttentionLayer.forward\_parallel} \\
\bottomrule
\end{tabularx}
\end{table}

The RWKV-6 and Mamba-2 LION deployments reuse the existing time-mix
and SSD block class with an added \texttt{mode} parameter; the
Linear Attention LION deployment uses a separate encoder class
\texttt{LIONLinearAttentionEncoder} distinct from the causal
\texttt{CausalLinearAttentionEncoder}. The decay-mode choice on the
LA deployment selects between LION-LIT and LION-S as described in
Section~\ref{sec:proposed:lion}.

\subsection{Decay-class mapping}

The natural decay-class mapping per~\citet{afzal2025linear}
Table~1 carries each causal backbone into LION mode while preserving
the causal backbone's decay class.
RWKV-6 causal $\to$ LION-S inherits the per-channel data-dependent
$w_{t}$. Mamba-2 causal $\to$ LION-S inherits the scalar per-token
data-dependent decay derived from the selective $\Delta t \cdot A$
parameterisation. Linear Attention causal $\to$ LION-LIT
inherits $\lambda = 1$, since the causal LA backbone is a pure
accumulator without intrinsic decay.

\subsection{LION-S falsification control on Linear Attention}

The Linear Attention backbone is additionally tested in a
controlled LION-S configuration in which a per-token sigmoid decay
is introduced that is not present in the underlying causal backbone.
This deployment is reported as a controlled experiment alongside
the LION-LIT default, since the LION-LIT vanilla cell on LA produces
test CER $0.2951$ (well below the causal LA test of $0.1879$) and
the LION-LIT $\times$ CVD cell produces an actively negative
transfer ($\Delta = +0.0243$). The LION-S configuration restores
decay-bounded behaviour to the LA bidirectional accumulator and
yields the productive transfer pattern documented in
Section~\ref{sec:lion-decay-prereq}; the contrast between the two
LA-LION deployments isolates the role of decay as a structural
prerequisite for the value-decorrelation mechanism on the
bidirectional substrate.
