\chapter{Conclusions}\label{ch:conclusions}

This thesis characterised three causal linear-time recurrent
backbones (RWKV-6, Mamba-2, Linear Attention) and three
mechanism-level extensions (Multi-Scale Depthwise Convolution,
Damped Harmonic Oscillator, Chunked Value Decorrelation) on a
cross-architecture transfer matrix at 7M and 14M parameter
scales. The matrix supports a predictive criterion that we call
mechanism-prior alignment: larger residual deficits along the
axis a mechanism targets tend to correspond to larger absolute
gains, and no measured gain is expected when the mechanism's
axis is not exercised by the task.

\section{Summary of findings}\label{sec:conclusions:summary}

The matrix populates the criterion in six structural signatures
across modes, scales, probes, and composition settings
(Section~\ref{sec:experiments:synthesis}): deficit-driven
ordering on MSDC, the LibriSpeech-only DHO direction (largest
absolute gain on Linear Attention, small absolute gains on
RWKV-6 and Mamba-2 under the audiobook distribution),
asymmetric transfer on CVD, the
LION-LIT~$\times$~CVD controlled falsifier isolating decay as
the bidirectional-mode prerequisite, the
Mamba-2~$\times$~DHO cross-distribution refinement to alignment
between mechanism and task-as-exercised-by-data, and the axis-2
isolation on the Multi-Query Associative Recall length sweep
where mechanism-augmented linear-time backbones solve the task
at $T = 1024$ while the causal Transformer fails. The matrix
ceiling on LibriSpeech is the 14M Mamba-2 CVD~$\times$~MSDC
composition cell at test character error rate $0.0618$; the
strongest single-mechanism gain is the LA LION-LIT~$\times$~DHO
BREAK at a $65\%$ relative reduction. The chapter does not
lead on these absolute or relative numbers; the artefact is the
matrix and the predictive criterion the matrix supports.

\section{Limitations}\label{sec:conclusions:limitations}

Four limitations bracket the empirical scope. First, the matrix
is single-seed for every reported cell. Per-cell standard errors
are not estimated. The across-matrix pattern is read off
approximately seventy statistical tests oriented in a single
coherent direction, which provides aggregate statistical
confidence that offsets the variance of any individual cell. The
single-seed methodology is a deliberate scope decision under the
available compute for the matched 50-epoch budget across 65
cells. Multi-seed validation of the high-magnitude cells, in
particular the LA causal DHO and LA LION-LIT DHO cells, would
sharpen the within-cell claim without changing the across-matrix
pattern. Second, the present matrix provides stronger evidence
for the positive-transfer side of the criterion than for the
full converse / null side; systematic null-effect validation is
left for future multi-seed work, and the criterion is reported as
a predictive heuristic supported primarily on its
positive-transfer half. Third, the matrix scales to 14M
parameters; the conditional 30M scale was dropped from scope
since no ranking inversion appears between 7M and 14M, but the
deficit-driven ordering is not asserted at scales beyond 14M.
Fourth, the acoustic probes are English-only; the Common Voice
cross-distribution validation exercises within-language acoustic
variability rather than cross-language transfer.

\section{Scope and future work}\label{sec:conclusions:future-work}

The criterion as stated holds within the diagonal-class function
family of Section~\ref{sec:bounds}. The matrix does not test it
on diagonal-plus-low-rank architectures, and the criterion is
not asserted on that class. The natural extension of the
framework is to repeat the cross-architecture transfer matrix
on a diagonal-plus-low-rank backbone class such as
DeltaNet~\cite{schlag2021deltanet, yang2024deltanet},
DeltaProduct~\cite{siems2025deltaproduct}, or
RWKV-7~\cite{peng2025rwkv7}, with mechanisms targeting axis~3
(non-Abelian state-tracking). The DeltaProduct parameterisation
is particularly suited as an axis-3 substrate, since its $n$-step
Householder product exposes a tunable expressivity dial inside
the diagonal-plus-low-rank class and allows the cross-architecture
transfer matrix to be populated at controlled state-tracking
capacity. The formal complexity-class boundary recalled in
Section~\ref{sec:bounds} brackets exactly where the diagonal class
cannot reach and where the diagonal-plus-low-rank class is the
principal route across.

A second extension addresses a structural epistemological point
about the criterion. The criterion as stated tests a mechanism's
gain against a post-hoc reading of what a task exercises along
each axis. A stricter falsifiability protocol would specify, for
each (architecture, mechanism, dataset) triple, a numerical
prediction made independently of the experimental result, derived
from dataset-level metrics that quantify the structural axes the
dataset exercises. Examples include a phoneme-to-syllable energy
spectrum for axis 1 on speech, a key-similarity entropy for axis
2 on synthetic recall, and analogous metrics for the remaining
axes. The thesis does not develop such metrics. Their development
is the natural prerequisite for converting the alignment
criterion from a predictive heuristic into a strictly falsifiable
claim.

A third extension direction is the cross-modality replication of
the matrix on vision and language probes. The thesis instantiated
MSDC on speech as one node in a five-paper cross-domain pattern
(Section~\ref{sec:related:cross-modality}). The cross-architecture
transfer matrix is portable to those modalities with appropriate
substitution of the axis-1 and axis-2 probes.

The framework remains agnostic to the specific evaluation domain
and to the specific mechanism instantiation. What it asserts is
the predictive structure: a mechanism's gain orders with the
residual structural deficit along the axis it targets, modulated
by what the task-as-exercised-by-data exercises. The empirical
matrix populates the criterion in three productive signatures
across the diagonal-class linear-time family and one controlled
falsifier; further work extends the criterion across the
diagonal-plus-low-rank boundary and across modalities other
than speech.
